---
title: "Module 7: Ring Width and Census model w/climate"
author: "Andria Dawson and Kelly Heilman"
format:
  html:
    code-fold: true
---

## Example of incorporating predictor variables directly into the state-space model estimation

## Outline

-   read in formatted climate data (or format here)
-   Run STAN model with climate regression
-   Visualize outputs
-   Plot biomass estimates


## Load packages to be used
We estimate the parameters of this state-space model in a program called `stan` which interfaces to `R` via a package called `rstan` (`cmdstan` is another package that is used.)
```{R}
library(rstan)
```


# load the model data
```{r}

```

# Stan Model
```{r}
# add R stan code here
```

# add the math notation of the model here:

## Running the model: Example for a short run
```{r}
# run a version with a small # of iterations here to show the output
```


## The model outputs
```{R}
# load the Full TR only model outputs
# put some example exploration of the model outputs here
```

## convert tree diameters to AGB
```{R}
# code to estimate biomass
```


## Plot of AGB and AGBI with uncertainty
```{R}
# plot total and by taxa
```
