---
title: "Module 3a: State-space model (RW-only)"
author: "Andria Dawson and Kelly Heilman"
format:
  html:
    code-fold: true
---

## State-space model for estimating biomass 

## Load packages to be used

We estimate the parameters of this state-space model in a program called `stan` which interfaces to `R` via a package called `rstan` (`cmdstan` is another package that is used.)

```{R}
library(rstan)
```

# load the model data

```{r}

```

# Stan Model

```{r}
# add R stan code here
```

# add the math notation of the model here:

## Running the model: Example for a short run

```{r}
# run a version with a small # of iterations here to show the output
```

## The model outputs


```{R}
# load the Full TR only model outputs
# put some example exploration of the model outputs here
```

## Model Checking in stan:

```{R}
# traceplots
# checking model diagnostics

```

## Model Validation:

```{r}
# comparing estimates of diameter increment to observations
```

## convert tree diameters to AGB

```{R}
# code to estimate biomass
```

## Plot of AGB and AGBI with uncertainty

```{R}
# plot total and by taxa
```
